Bacon: A GPU Programming Language With Just in Time Specialization

### 2011-05-03 doc/Paper/Paper.tex
Tests pass again.

This paper describes Bacon, a programming language that makes it
easier to write high performance kernels to run on for OpenCL
compatible general purpose GPUs. The design and implementation is
described, and benchmarks are given showing performance gains over
comparable OpenCL programs.

@@

The OpenCL standard consists of two major pieces. First, it defines a
programming language called OpenCL C for writing compute kernels to
run on parallel hardware. Second, it defines runtime APIs for C and
C++ that allow these kernels to be compiled, loaded, and executed from
programs running on a host machine.

@@

The key optimization performed by the Bacon compiler is just in time
specialization. When a kernel is called, a specialzied version of the
kernel is generated with some of its variable arguments replaced with
compile-time constants. This allows for a variety of agressive
optimizatons to be performed by the OpenCL compiler. Further, it
allows for some variable sized array declarations to be turned into
constant sized arrays that are allowed by OpenCL.

@@

@@

The system is built using Perl and C++. The front-end compiler parses
the source code using Parse::Yapp, a yacc-compatible parser generator
for Perl. This constructs abstract syntax tree, which is then walked
to generate the C++ wrapper. The abstract syntax tree (AST) is then
serialized for later loading by the runtime library.

@@

In order to evaluate the performance of the Bacon system, we compare
the run time of matrix multiplication kernels in both Bacon C and
OpenCL C.

Testing was performed on an ATI Radeon 5830 GPU. This is a cut down
version of AMD's current top of the line GPU chipset, which at the
time of this writing sells new for around \$140. A similar Nvidia card
was purchased for comparison testing, but the comparsion was not
performed due to implementation incompatibilities that we have not yet
resolved.

@@

Tests were performed on 4096x4096 matrices to create large enough
chunks of work to be able to time the results without worrying about
factors like OpenCL queue latencies. As table~\ref{mm1} shows, the
optimizations provided by the Bacon system provide significant
speedups on this test. As a comparison to show how good GPUs are for
this kind of taask, a naive serial implementation running on a CPU can
take as much as 45 minutes to do this calculation\footnote{Tested on
  an AMD Phenom II X3 @ 2.8 GHz}, although a highly optimized
implementation can approach the GPU times\footnote{ATLAS BLAS clocks
  in at 7.8 seconds}.

@@

\section{Conclusion}
Hai \cite{Berlin:1990}.

>>>>

This paper describes Bacon, a programming language that makes it
easier to write high performance kernels to run on for OpenCL
compatible general purpose GPUs. The design and implementation is
described, and benchmarks are given showing performance gains over
comparable programs written directly in OpenCL C.

@@

The OpenCL standard\cite{opencl} consists of two major pieces. First,
it defines a programming language called OpenCL C for writing compute
kernels to run on parallel hardware. Second, it defines runtime APIs
for C and C++ that allow these kernels to be compiled, loaded, and
executed from programs running on a host machine.

@@

The key optimization performed by the Bacon compiler is just in time
specialization. Rather than performing When a kernel is called, a
specialzied version of the kernel is generated with some of its
variable arguments replaced with compile-time constants. This allows
for a variety of agressive optimizatons to be performed by the OpenCL
compiler. Further, it allows for some variable sized array
declarations to be turned into constant sized arrays that are allowed
by OpenCL.

@@

\section{Previous Work}

Specialization, also known as partial evaluation, has been shown to be
a very effective optimization by projects such as
Tempo\cite{consel:1998}, which does general partial evaluation of C
code at compile time. Just in time specialization was popularized by
its use in typing dynamic languages, as shown in the Self
language\cite{chambers:1992}. Just in time specialization on values has
been used in Prolog systems by Bolz\cite{bolz:2010}.

@@

The system is built using Perl and C++. The front-end compiler parses
the source code using Parse::Yapp\cite{parse::yapp}, a yacc-compatible
parser generator for Perl. This constructs abstract syntax tree, which
is then walked to generate the C++ wrapper. The abstract syntax tree
(AST) is then serialized for later loading by the runtime library.

@@

In order to evaluate the performance of the Bacon system, we compare
the run time of matrix multiplication kernels in both Bacon C and
OpenCL C. These results are summarized in Table~\ref{mm1}, which shows
that Bacon is able to provide measurable performance improvements over
similar programs written directly in OpenCL C.

Testing was performed on an ATI Radeon HD 5830 GPU. This is a cut down
version of AMD's current top of the line GPU, the Radeon HD 6970. The
5830 sells for around \$140 at the time of this writing compared to
\$340 for the 6970, while providing over 60\% of the theoretical
compute power. A similar Nvidia card was purchased for comparison
testing, but the comparsion was not performed due to implementation
incompatibilities that we have not yet resolved.

@@

Timings were taken multiplying 4096x4096 matrices. Each test was
performed five times and the average result was taken. The times were
very consistent; most tests had a coefficient of variation under one
percent. The speedups over the naive OpenCL implementationa are shown
in Figure~\ref{unroll}.

@@

These measurements clearly show that value specialization provides
significant speedups over a non-specialized OpenCL kernel. This result
was expected; providing constant values at compile time allows the
OpenCL C compiler to do extensive and agressive constant
propagantion-based optimizations.

The results from loop unrolling are more complicated. Unrolling loops
too much drastically decrease performance, most likely due to the
exhaustion of registers on the GPU device. Still, when properly tuned,
loop unrolling provides significant speedups allowing the blocked
Bacon kernel to beat the hand vectorized OpenCL kernel by nearly 30
percent.

\section{Future Work}

The Bacon system could benefit from a number of extensions to improve
its performance, generality, and utility to developers. One obvious
area of improvement would be to automatically determine the best
factor for loop unrolling rather than expecting it to be tuned
manually. The unrolling analysis could also be extended to the
implicit parallel ``loops'' created by the range of the computation;
optimally, it should be possible to generate kernels like the blocked
matrix mulitplication shown from the naive matrix multiplication
kernel. Another simple improvement would be to perform strip mining
(Wolfe \cite{wolfe:1996}, section 9.8) and generate explicit vector
code for loops.

Just in time specialization as a method to improve performance on
parallel computations seems very promising in general, but there are
some limitations imposed by the use of OpenCL as a compiler target. An
implementation of this technique targetting a parallel processor at a
lower level would allow for more flexibiltiy and much faster
specialization times.

\section{Conclusion}

We have shown that Bacon allows a high performance GPU matrix
multiplication kernel to be written in a naive style and executed with
nearly the performance of a hand-vectorized OpenCL kernel due to the
performance benefits of just in time specialization. Further, we have
shown that by re-writing that kernel with a generalized block strategy
and selecting appropriate loop unrolling settings, the performance of
the hand vectorized OpenCL kernel can be exceeded.

We are distributing this tool in the hope that it will be practically
useful for developing kernels targeting OpenCL compatible GPU devices.
